Robertson and Cao Project

### 2024-04-17 code/analysis-code/analysis2.qmd
Updated manuscript analysis

#### Loading library and data set
``` {R}
# Load required libraries
library(MASS) # for robust regression
library(caret) # for cross-validation
library(h2o) #For dimension reduction
library(dplyr)

@@

@@

# Convert mydata to H2OFrame
mydata_h2o <- as.h2o(mydata)

@@

>>>>

#### Loading library and data set
``` {R, warnings=FALSE, messages=FALSE}
# Load required libraries
library(MASS) # for robust regression
library(caret) # for cross-validation
library(h2o) #For dimension reduction
library(dplyr) 
library(rsample)

@@

# Set Seed for reproducibility
rngseed = 1234
set.seed(rngseed)

# Create training data by alloting 3/4 of the data for training
data_split <- initial_split(mydata, prop = 3/4)

# Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)

@@

# Convert train_data to H2OFrame
mydata_h2o <- as.h2o(train_data)

@@

```{r}
# Replicating the previous analysis with test data

# Convert train_data to H2OFrame
mydata_h2o <- as.h2o(test_data)

# Create a fold column for cross-validation
fold_column <- h2o.kfold_column(data = mydata_h2o, nfolds = 10, seed = 1)  # Specify the number of folds and a seed for reproducibility

# Display results
print(cv_metrics)

### 2024-04-17 products/manuscript/Manuscript.qmd
Updated manuscript analysis

# Methods
## Loading Data
The data was loaded from the original DAT file, entitled 'NISTEENPUF22.DAT' by specifying the column positions within the data file. The columns were then read and made into a data frame. Each of the factor variables was defined and the corresponding levels were assigned labels. The structure and summary statistics of all of the variables were examined. The loaded data is saved as a rds and a csv in the 'processed-data' folder, as to not alter the original raw data file. The definition of each factor variable in included in the README file within the 'raw-data' folder.

## Cleaning Data
The structure of the processed data was examined.  The data was examined for missing values. No NAs were present, however missing values were present as labeled factor levels. The missing factor levels were printed, identified, and replaced with NA values. A string of 27,000 NAs were present in the outcome variable, indicating vaccination completion. All rows corresponding to the NAs for the missing outcome value were removed. Approximately 160 additional rows with NA values for facility or insurance status variables were removed. Over 16,000 rows of data remained. The year column was eliminated, as all of the data is from 2022. We saved the cleaned data as an rds and csv in the 'processed-data' folder.
## Exploratory Analysis
Exploratory analysis was used to determine the distribution of the predictors, including the socioeconomic variables that were chosen for analysis. The distribution of single predictors including income-to-poverty ratio, income, state, race, ethnicity, vaccination status, and facility were examined using histograms and scatter plots. The following predictors were compared visually and fitted to a linear regression to examine preexisting associations: vaccination status and ethnicity, vaccination status and poverty status, income poverty ratio and insurance status, income poverty ratio and ethnicity. 
## Analysis
### Generalized Logistic Regression
A generalized logistic regression was performed with bi-directional step-wise selection for predictor reduction. Variables that were redundant of a major variable were excluded from analysis. The variables initially included were age, sex, state, race/ethnicity, insurance status, income, insurance breaks, poverty status, education, language, geographic mobility, facility, and housing status. The model was evaluated with a residual plot and correlation matrix to reveal strong multicollinearity. This is the basis for the subsequent analysis methods, as they are more resistant to collinearity.
### Principal Component Analysis
Principal component analysis (PCA) was performed to do dimensionality reduction. PCA would ideally transform the highly correlated predictors into uncorrelated components. The h2o package was used to provide a machine-learning platform in RStudio, to allow machine-learning methods ot be implemented. Due to the data being mainly categorical variables, the PCA proved to be unsuccessful. A LASSO Ridge Regression was performed with the h2o package, to mitigate the multicollinearity issue.
### Random Forest Model
A random forest model was used, as it is resistant to multicollinearity. The tidymodels package was used to produce a random forest model with the classification method. The model was trained and fit from training data, which was an 80% split of the original data. After evaluation, the model parameters for minimal node size and number of selected predictors were tuned for using a grid search with cross-validation resamples. After tuning, the best model was extracted based on the ROC AUC metric and was fit to the training data. The tuned model was used to make predictions and evaluated with performance metrics, a confusion matrix, and a comparison to a single-predictor random forest model. Finally, the tuned random forest model was re-evaluated by being fit to the testing data.
### LASSO Regression

@@

# Results
## Exploratory Data Analysis
Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis. Nearly all of the demographic factors have a similar distribution between those who have completed their HPV vaccination regime versus those who have not completed it.

```{r}
#| label: Table1
#| tbl-cap: "Table 1.Socioeconomic characteristics of U.S. teenagers who have completed or are up-to-date with HPV vaccination series."
#| echo: FALSE
Table1 <- readRDS(here("results/tables/table1.rds"))
kable(Table1)
```
The distribution of the numerical income to poverty ratio can be shown below.
It displays a value from 1-3.
A ratio displaying less than 1 depicts an income less than the poverty level.Any income ratio \>1 indicates an income greater than the poverty level.
Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.

@@

We begin with various socioeconomic predictors that interest us.
Some of these predictors are a subset of the other, or may be used to describe an overall category.
For this reason

@@

![Table 3. Human Papillomavirus vaccine status variables and meanings.](assets/UTDHPV-variable-table.png)

@@

![Table 4. Potential predictor variable names and meanings.](assets/Predictor-variable-table.png)

@@

To determine which predictor variables will be best suited for a general linear model, I will do a step-wise comparison and then measure the resulting models with AIC and WAIC.

# Discussion

>>>>

# Methods

## Loading Data

The data was loaded from the original DAT file, entitled 'NISTEENPUF22.DAT' by specifying the column positions within the data file.
The columns were then read and made into a data frame.
Each of the factor variables was defined and the corresponding levels were assigned labels.
The structure and summary statistics of all of the variables were examined.
The loaded data is saved as a rds and a csv in the 'processed-data' folder, as to not alter the original raw data file.
The definition of each factor variable in included in the README file within the 'raw-data' folder.

## Cleaning Data

The structure of the processed data was examined.
The data was examined for missing values.
No NAs were present, however missing values were present as labeled factor levels.
The missing factor levels were printed, identified, and replaced with NA values.
A string of 27,000 NAs were present in the outcome variable, indicating vaccination completion.
All rows corresponding to the NAs for the missing outcome value were removed.
Approximately 160 additional rows with NA values for facility or insurance status variables were removed.
Over 16,000 rows of data remained.
The year column was eliminated, as all of the data is from 2022.
We saved the cleaned data as an rds and csv in the 'processed-data' folder.
## Exploratory Analysis Exploratory analysis was used to determine the distribution of the predictors, including the socioeconomic variables that were chosen for analysis.
The distribution of single predictors including income-to-poverty ratio, income, state, race, ethnicity, vaccination status, and facility were examined using histograms and scatter plots.
The following predictors were compared visually and fitted to a linear regression to examine preexisting associations: vaccination status and ethnicity, vaccination status and poverty status, income poverty ratio and insurance status, income poverty ratio and ethnicity.
## Analysis \### Generalized Logistic Regression A generalized logistic regression was performed with bi-directional step-wise selection for predictor reduction.
Variables that were redundant of a major variable were excluded from analysis.
The variables initially included were age, sex, state, race/ethnicity, insurance status, income, insurance breaks, poverty status, education, language, geographic mobility, facility, and housing status.
The model was evaluated with a residual plot and correlation matrix to reveal strong multicollinearity.
This is the basis for the subsequent analysis methods, as they are more resistant to collinearity.
### Principal Component Analysis Principal component analysis (PCA) was performed to do dimensionality reduction.
PCA would ideally transform the highly correlated predictors into uncorrelated components.
The h2o package was used to provide a machine-learning platform in RStudio, to allow machine-learning methods ot be implemented.
Due to the data being mainly categorical variables, the PCA proved to be unsuccessful.
A LASSO Ridge Regression was performed with the h2o package, to mitigate the multicollinearity issue.
### Random Forest Model A random forest model was used, as it is resistant to multicollinearity.
The tidymodels package was used to produce a random forest model with the classification method.
The model was trained and fit from training data, which was an 80% split of the original data.
After evaluation, the model parameters for minimal node size and number of selected predictors were tuned for using a grid search with cross-validation resamples.
After tuning, the best model was extracted based on the ROC AUC metric and was fit to the training data.
The tuned model was used to make predictions and evaluated with performance metrics, a confusion matrix, and a comparison to a single-predictor random forest model.
Finally, the tuned random forest model was re-evaluated by being fit to the testing data.
### LASSO Regression

@@

# Results

## Exploratory Data Analysis

Exploratory analysis is used to determine the distribution of our predictors, including the socioeconomic variables that we chose for analysis.
Nearly all of the demographic factors have a similar distribution between those who have completed their HPV vaccination regime versus those who have not completed it.

```{r}
#| label: Table1
#| tbl-cap: "Table 1.Socioeconomic characteristics of U.S. teenagers who have completed or are up-to-date with HPV vaccination series."
#| echo: FALSE
Table1 <- readRDS(here("results/tables/table1.rds"))
kable(Table1)
```

The distribution of the numerical income to poverty ratio can be shown below.
It displays a value from 1-3.
A ratio displaying less than 1 depicts an income less than the poverty level.Any income ratio \>1 indicates an income greater than the poverty level.
Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.

@@

We begin with various socioeconomic predictors that interest us.
Some of these predictors are a subset of the other, or may be used to describe an overall category.

@@

![Table 3. Human Papillomavirus vaccine status variables and meanings.](./images/UTDHPV-variable-table.png)

@@

![Table 4. Potential predictor variable names and meanings.](./images/Predictor-variable-table.png)

@@

To determine which predictor variables will be best suited for a general linear model, we employed a stepwise comparison is performed and then assessed with AIC and WAIC. The stepwise model is set up with `P_UTDPHPV` as the outcome and the remaining chosen variables as the predictors. Each predictors were added and dropped in different variations, and the resulting AIC were recorded for each of the different models. From this, we were able to determine that the model with the lowest AIC (and thus the greatest fit) was the model with all of the predictors that we previously picked out.

Moving from that, we plotted the residuals from the logistic regression model to check for non-linearity and discovered that the predictors potentially had a high indication of collinearity. to further confirm this, a VIF analysis was attempted but was unable to proceed as the multicollinearity was too great for too many variables. As a result, we decided to perform a correlation matrix, which performed as expected and verified our concerns.

INCLUDE EITHER REGRESSION PLOT OR STEPWISE FINAL RESULT IMAGE

In order to reduce the multicollinearity issue, we initially attempted to use PCA. To employ the unsupervised learning model, we use the `h2o` package. The PCA resulted in one principle component with a high standard deviation and high proportion of variance, showing that the Pc1 captures a large portion (97.9%) of variability in the data.

INCLUDE IMAGE OF PCA RESULTS

The result of having only one principle component further supports the fact that many of the predictor variables are highly collinear. One PC shows that a PCA must reduce the data object to a single dimension. Having a single principal component could be problematic for my analysis, particularly if it leads to the loss of crucial information or structure in my data due to collinearity among the original features.

Using the same package, I used the LASSO logistical regression model and performed a co-validation test on it.

```
MSE:  0.1558815
RMSE:  0.3948183
LogLoss:  0.4850906
Mean Per-Class Error:  0.4906615
AUC:  0.6596287
AUCPR:  0.8704217
Gini:  0.3192575
R^2:  0.05771195
Residual Deviance:  16070.08
AIC:  16266.08
```

Random Forest

LASSO Regression
# Discussion
